% !TEX TS-program = xelatex
% !TEX encoding = UTF-8

\documentclass[a4paper, 11pt]{article} % use larger type; default would be 10pt

\usepackage{dfdocs}

\usepackage{xr}
\externaldocument[const.]{df_constitution}

\title{DF Democratic Procedures}
\author{DF Secretary}
\date{\today}

\begin{document}
\maketitle

\tableofcontents

\section{Foreword}
Where the procedures specified in this document conflict with the DF Constitution, then the procedures in the Constitution shall prevail.  That is to say, this document may not override the Constitution.

\section{Business}
In all DF business (for example, at Althing, a Thing, or a mini-Thing), a system of consensus-based decision making should be used.  Participants are strongly encouraged to read ``The A--Z of Good Discussion'', which gives a good outline of the consensus system.  Recommendations for consensus are also covered in \autoref{sec:consensus} of this document.

Additionally, at the beginning of business, the chair will briefly explain the principles and practicalities of this system, in particular the hand signals being used.  The chair may of course also deliver this explanation at later points in the discussion if, for example, new people have joined in or on return from a break.

\subsection{Quoracy}
No business may be undertaken without a quorum of members present, as outlined in the Quorum section of the Constitution.

\subsection{Timing of Business}
All business will take place between 8am and midnight, and no business will take place over a mealtime.

\subsection{Expenses to Business Events}
\subsubsection{Things}
The following people expected to come to Things and therefore can be exempt from paying the full event cost of the Thing and claim travel expenses from the Treasurer:
\begin{itemize}
\item all members of DF Committee
\item 2 members of the organising team for the next social event
\item three volunteers in a non-expensed role who would benefit DFs by attending the Thing
\item the two DFs elected to General Council from the floor of Althing
\item the Accessibility Representative
\end{itemize}

Additionally, in the year running up to a Venturer Camp, the role of Venturer Committee Liaison is also mandated and expensed to come to Things.

The three non-specific volunteer places are to be applied for to DF Committee.

\subsubsection{Althing}
The following people are expected to come to Althing and can be exempt from paying the full event cost of Althing and may claim travel expenses from the Treasurer:
\begin{itemize}
\item all members of DF Committee
\item The Accessibility Representative
\item Venturer Committee liaison
\end{itemize}

\section{Motions}
A motion is a specific instruction or statement agreed upon as reflecting the wishes or opinions of the DF Movement at large.  Motions are debated and voted on formally at Althing, and it is the job of the Secretary to prepare and publish an agenda of motions in advance of Althing.  All motions must have a proposer, who submits the motion for debate at Althing. If a motion is agreed upon, it is said to \emph{pass} or \emph{be carried}, and if not, it is said to \emph{fall}.

\subsection{Submission}
A motion may be submitted by any DF or DF Committee.  Motions may be submitted at any point up to and including Althing, however the Secretary may set a deadline by which motions must be submitted in order to be included on the agenda.  Motions submitted after this deadline will only be discussed at Althing if there is enough time after all previously submitted motions.

Motions may be `seconded' by a DF or DF Committee.  The seconder of a motion must have to be distinct from its proposer.

\subsection{Discussion}
When it is a turn of a particular motion to be discussed, the chair will call for somebody to speak for the motion, usually the proposer.  If nobody wishes to speak for the motion, then it automatically falls and business moves on to the next agenda item.  If somebody is found to speak for the motion, the chair then immediately calls for somebody to speak against it.  If nobody is found to speak against it, then the motion is automatically carried.  If people are found to speak both for and against it, then these two people open the debate.  Once they have finished, the motion is discussed by everybody using the usual consensus system.

If a discussion concludes naturally, or appears to be going round in circles with no fresh ideas, or needs to end because of time constraints, the chair may take several more points and then move to a vote.  Immediately before a vote, the proposer of the motion has the right to make a final point, known as the \emph{right of reply}.

\subsection{Voting}
\label{sec:motionvoting}
Participants may cast a vote for a motion, or against it, or they may abstain from voting.  Voting on motions is not usually done in secret, however the discussion as a whole may decide (by consensus or otherwise) to hold the vote in secret.  For motions which do not alter the Constitution to pass, a simple majority is required, i.e. 50\% + 1 vote.  Motions which do alter the Constitution require a 2/3rds majority vote to pass, i.e. 2/3rds + 1 vote.  When counting totals of votes cast, abstentions do not count, and quorum applies only to those votes cast.

\subsection{Amendments}
At any time up until voting on a motion, amendments may be submitted (by a DF or DF Committee) which change an aspect of the motion.  Discussion and voting for amendments always takes place before discussion of the motion, and then the motion is discussed `as amended' or otherwise.

The format for discussing and voting on amendments is the same as for motions, with the following exceptions.
\begin{itemize}
\item You cannot propose an amendment to an amendment.
\item The proposer of a motion may choose accept an amendment, in which case the amendment is applied automatically without discussion or a vote.
\item All amendments not accepted by the proposer of the motion are decided by a simple majority vote.
\end{itemize}

\subsection{Conflicting Motions/Amendments}
\label{sec:conflicting}
Where two or more motions or amendments conflict, i.e. negate one another fully or in part, the secretary, with the help of DF Committee, will work out the best way to discuss the motions when writing the Althing agenda.

\section{Musings}
A musing consists of a thought or a question, formulated to inspire creative discussion.  A musing is less formal than a motion, and are discussed in the `open space' system, guidelines for which are to be found in \autoref{sec:openspace} of this document.  Musings have an open result; that is to say the actions to come out of a musing are not predefined.  It may be the case that a motion is formulated and proposed as a result of a musing, but equally a working group may be set up, or a project started by interested parties.

\section{Policies}
A Policy is a document that is used to describe the procedures of DFs. These Policies are not required to be put in place through a motion at Althing and must follow the existing procedures within the Constitution and Democratic Proceedings, in the case of a conflict between a Policy and the Constitution or Democratic Proceedings, then the procedures in the Constitution and Democratic Proceedings shall prevail.

In order for a new Policy or change to an existing Policy to be accepted by the DF movement, a vote must be held at a Thing. This vote must require a simple majority after the attendants of the Thing are able to discuss and amend the Policy where appropriate. In order for this vote to carry the attendance of the Thing should include at least 5 members in which at least 1 committee and 1 non-committee member are present. Any new Policies or Policy changes should be reviewed within 2 years.

Policies and Policy changes can be put forward by the following groups:
\begin{itemize}
\item The DF Movement
\item DF Committee
\item General Council
\item Venturer Committee
\end{itemize}

\section{Elections and Roles}
Where the election for a particular role is held depends on the motion which created the role, however, most elections take place at Althing.  In all elections, a virtual candidate called `RON' (standing for Re-Open Nominations) shall be electable.  In most cases, if RON is elected, a request for further nominations of candidates is put out, and the election is postponed until the next available business event.  The exception to this, as detailed in section \ref{const.sec:treasurernoconfidence} of the Constitution, is the role of Treasurer.

\subsection{Nominations}
Before an election is held, nominations for candidates may be made to the Secretary.  The Secretary will establish if each nomination is accepted by its nominee, and if so their name will be included in the final list of candidates.

The virtual candidate `Re-Open Nominations' is automatically nominated in all elections.

\subsection{Hustings and Voting}
At the beginning of the election, the Secretary will close nominations and invite all the candidates to give a short speech, known as a \emph{hustings}.  The purpose of the hustings is to allow candidates to inform and persuade voters of their electability.  A candidate does not have to give a hustings, but if they do it is recommended that they take questions at the end of it.

All elections to single roles shall take place by AV.  Elections where there are multiple positions available to the same pool of candidates will take place by STV.  This includes, for example, the special case where one Lay Member stands down mid-term, and therefore both Lay Members are elected in the same year.

\section{Voting Practicalities}
Voting is usually required to pass a motion, and to elect a candidate.  As detailed in section \ref{sec:motionvoting}, voting on motions is usually done by a simple show of hands for votes for, votes against and abstentions.  However, elections for roles and certain special cases for motions (conflicting motions or amendments, see \ref{sec:conflicting}) use voting systems that require some explanation.

\subsection{Secret Ballots}
Elections are usually held in secret, with votes expressed written down anonymously on ballot papers.  These votes are then collected and counted in public, and the result announced to the electorate.  When voting on motions is held in secret, a similar process may be used, or possibly for convenience, members of the discussion may be asked to close their eyes and vote normally with hands.

\subsection{Tellers}
This of course raises the question of who does the counting.  These people are known as \emph{tellers} and are chosen by consensus from a discussion.  It is their job to make sure voting is carried out and votes are counted correctly.  To make sure tellers are (relatively) neutral, they must surrender their right to vote in the election they are administering.

\subsection{Voting Systems}
The type of voting system used in an election will depend on the details of the election.  Usually, AV is used for elections with one position available, STV for elections with multiple positions and Run-off Voting used as an alternative to AV, when AV would be too time consuming.

\subsubsection{Alternative Vote (AV)}
\label{sec:av}
Under the Alternative Vote system, a single candidate is elected from a pool of candidates.  Each voter expresses a ranked list of preferences on their ballot paper, in the order in which they would prefer those candidates to win.  Then, once collected, the following process is used to establish a winner of the election:

\begin{enumerate}
\item Votes are counted based on first preferences, discarding spoilt votes.
\item \label{item:candidatemajority} If a candidate has a 50\% + 1 majority, they are elected.
\item Otherwise, the candidate with the least votes is eliminated, and their votes are redistributed based on the next preference.
\item Return to step \ref{item:candidatemajority}.
\item If there is a tie, the election is held again with only those two candidates eligible for votes.
\item If there is still a tie, a coin-toss may decide the vote, or the two tied candidates may agree to share the position in some way.
\end{enumerate}

\subsubsection{Run-off Voting}
This system is functionally equivalent to AV, in that if all voters have the same preferences, the outcome of the election will be the same.  It is generally used when running an AV election would be too time consuming or otherwise impractical.  A disadvantage is that it cannot be run as a secret ballot.

Each candidate is assigned a physical space, e.g. a particular corner of the room, and voters are then asked to move to the space of their preferred candidate.  Then, the candidate with the least number of people in their space is eliminated, and those people are asked to move to the space of their second-most preferred candidate.  The process is repeated, as in AV, until one candidate has a 50\% + 1 majority.

\subsubsection{Single Transferrable Vote (STV)}
Single Transferrable Vote is a system by which candidates are elected to multiple positions achieving proportional representation.  When there is only one position available, it is equivalent to AV.  As in AV, each voter expresses a ranked list of preferences on their ballot paper and the following process is used to decide the winners:

\begin{enumerate}
\item The quota of votes for a candidate to be elected is decided, usually by the following formula (rounded up)\footnote{This ensures that the percentage of votes held by those elected cannot be less than or equal to the votes held by anybody not elected.} \[\mbox{votes needed to win} = \frac{\mbox{votes cast}}{\mbox{positions to fill}+ 1} \]
\item Votes are counted based on first preferences, discarding spoilt votes.
\item \label{item:stvmajority} If a candidate reaches (or surpasses) the quota, they are elected.
\item Any votes for an already-elected candidate are redistributed to their next preference.\footnote{This minimises votes `wasted' on candidates who have already been elected.}
\item If all votes have been counted, the candidate with the fewest votes is eliminated and their votes are redistributed based on the next preference.
\item Return to step \ref{item:stvmajority}.
\end{enumerate}

\appendix

\section{Open Space}
\label{sec:openspace}
This section sets out the practicalities of the `Open Space' system of discussion.  It is a system designed to encorage creativity and co-operation in solutions to problems.

Items are discussed simultaneously by small groups of people, who then feed back into the main discussion afterwards.  There are 3 principles, 3 (sadly metaphorical) animals, and a law which dictate the organisation of these small discussions.

The outcomes of the discussions are by no means constrained.  They could be (for example) a proposal, an action, a collection of more ideas, or a question.

\subsection{The Law of Movement}
The Law of Movement says that you are free to leave or join any dicussion at any time.  You should not feel obliged to stay in one, nor should you feel obliged to join another.  You and only you are responsible for deciding which discussions to take part in.

\subsection{Principles}
\begin{description}
\item[The right people for the discussion are whoever comes.]  Don't think about who might be best to participate in the discussion; be happy with the people there.
\item[Whatever happens is all that could have happened.]  There's no point worrying that something important wasn't said.  If it wasn't said, it probably wasn't important.
\item[When it's over, it's over.]  It's OK if a discussion reaches a natural conclusion.  Similarly, any kind of outcome is OK, whether it's concise, conflicting, or even another question.
\end{description}

\subsection{Animals}
\subsubsection{The Giraffe}
Giraffes often find themselves using their long necks to listen in on other discussions that they might find more interesting, and as a result they don't participate fully in their current discussion.  If you find yourself being a giraffe, use the Law of Movement to move to the other discussion.

\subsubsection{The Butterfly}
Butterflies sit and look pretty, often just having a chat over a hot drink and maybe a biscuit.  They don't talk about things related to the discussions going on, but sitting and looking pretty is useful too.

\subsubsection{The Bee}
Bees buzz about from group to group, cross-pollinating ideas and thoughts as they go.  They bring variety and richness to the discussions they land at.

\section{Consensus}
\label{sec:consensus}
In this section, some recommendations will be made regarding consensus-based decision making.  It is important, however, to note that this document recognises that this is a living system and therefore subject to change.  As such, it should be recognised that these are only recommendations, and if a particular business event wishes to adapt or modify them for its own purposes then it should feel free to do so.

\subsection{Chairing and speaking}


Chairing should involve maintaining a list of people who wish to speak, firstly so that you can keep track of what order people raised their hands in, and secondly so that they don't have to keep their hands in the air until it is their turn to speak.  It is also good for the chair to periodically sum up the discussion and provide insight into proposals.  (Note that this does not mean adding more points or ideas to the discussion, merely summarising what has already been said.)

Business should begin naturally with one person chairing and introducing the topics for discussion, and they may choose to continue chairing the discussion throughout.  However it is important that the participants feel able to challenge the authority of the chair for a good reason (for example if the chair is biased or abusing their authority), and if they choose, appoint a new chair.  It is also good practice for different people to chair the discussion at different points so that more people can gain experience of chairing discussions.

\subsection{Important distinctions: Proposals, Points, and Direct Points}
\subsubsection{Proposals}
If you have a good idea of something to do, i.e. a specific action, you should make a proposal.  Proposals `jump the queue' and should be discussed immediately.  After a decision is made whether to carry out the proposal, the order of speakers should return to that which it was before the proposal was made.

For example:
\begin{itemize}
\item ``Why don't we set up a working group to look into this?''
\item ``I have a friend with a printing press, shall I ask her if we can use it?''
\item ``We should have this camp at XYZ campsite!''
\end{itemize}

\subsubsection{Points}
To participate in a discussion, you should indicate (to the chair) that you want to make a point, usually by raising your hand.  The chair will then call on you at a suitable time, when you should make your point in a clear and accessible manner.  You should make sure to listen to other people's points as well as making your own, and if your point has already been covered by another person, don't say it again.  A point can be as long or as short as you like, within reason, although if the discussion is pressed for time, the chair may ask you to wrap your point up.

\subsubsection{Direct Points}
If you have a direct response to something somebody has said, you may use the `direct point' hand signal (usually both hands raised), and jump the queue of speakers. Direct points must be used with caution and should be limited to a few sentences at most. They are points of information, and should never express an opinion.

For example:
\begin{itemize}
\item (In response to, ``How many people were at last Venturer Camp?'') ``734.''
\item (In response to, ``Maybe we should use Lockerbrook for this event.'') ``Lockerbrook is fully booked until November.''
\item (In response to, ``I think we should set up a DF Libarary.'') ``We are already in the process of doing exactly that, talk to Esther Price or Joe MacMahon if you want to get involved!''
\end{itemize}

\subsection{Technical points}
A technical point is an important point about something unrelated to the discussion.  It is usually something practical and for this reason technical points jump the queue of speakers.

For example:
\begin{itemize}
\item ``The train leaves in half an hour; we need to hurry up.''
\item ``This building isn't wheelchair accessible, can we move elsewhere?''
\item ``I'm going to make tea, would anybody like some?''
\item ``Where are the toilets?''
\item ``It's getting far too late to have a reasonable discussion.  I think we should go to bed.''
\end{itemize}

\end{document}
